% Data 

## Raw data 

Historical weather data from a Met Éireann Oak Park synoptic weather station was used for model evaluation. The trial sites were in the radius of up to 500 m from the station in all years. 

## Data wrangling

```{r message=FALSE, warning=FALSE}
library("tidyverse")
#Weather data, parameters and cut off dates
files <- list.files("./data/", recursive = T, pattern = "Birr|Cas|Kil")
df_list <- lapply(files, function(x)  
  df_fun <- read_csv(paste0("./data/", x))
)

#Load&merge data for Casement station
df<-read.table("./data/hlydata2000_3723.txt", 
                     header=T,
                     stringsAsFactors=F, 
                     sep = ",")
glimpse(df)

```
The original files were modified as following. The names of the variables were changed to.. The original wide format was retained and will be transformed using the dplyr functions, etc.

## Glimpse 

```{r message=FALSE, warning=FALSE}
#Weather data, parameters and cut off dates

```

## Description of variables

- Variable1: 
  - Description 
  - Type: character
  - Unit

- Variable2: 
  - Description 
  - Type: integer
  - Unit

- Variable3: 
  - Description 
  - Type: double
  - Unit:
  
## Quick view 

```{r}
df
```


## Download 

The data are deposited in xx public repositories where they can be downloaded. Check on each data repository the policies regarding data use and citation.

<img src = "https://osf.io/static/img/circle_logo.png" width = "25">
[](DOI 10.17605/OSF.IO/7D3CH) 



